---
title: "Yet One More Plot to Show US's Race Inequality!"
author: Arafath Hossain
date: '2021-02-13'
slug: []
draft: true
categories:
  - R
tags:
  - Data Viz
  - TidyTuesday
subtitle: 'A short tutorial that shows the growing difference in student debt among the major three races in the US.'
description: 'Animated plot is a great way to show a trend. In this tutorial I looked into the students debt data that came from TidyTuesday. Using an animated line graph I will show how student loan has grown disproportionately among the three major three races in the US.'
image: 'https://images.unsplash.com/photo-1610316881910-0aac60da16b2?ixid=MXwxMjA3fDB8MHxwaG90by1wYWdlfHx8fGVufDB8fHw%3D&ixlib=rb-1.2.1&auto=format&fit=crop&w=1050&q=80'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Background

This week's [TidyTuesday](https://github.com/rfordatascience/tidytuesday/) was all about wealth inequality in the United States. Since I work for a university, I have seen how students' success varies among races. Looking back at their socio-economic backgrounds makes it quite obvious that there is a widening wealth gap among different races. Which makes the student debt data an interesting case for me to look at.

So here's the outcome of that. An animated graph that will show how America's races have historically been different in their student debt accumulation and how it's getting worse increasingly.

## Data

As said earlier, the data comes from TidyTuesday. Here's the [detail](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-02-09/readme.md) about the data.

```{r library_&_data, message=FALSE, warning=FALSE}
# libraries
# devtools::install_github("thomasp85/transformr")
library(ggplot2)
library(tidyverse)
library(ggtext)
library(gganimate) 
library(extrafont)
library(knitr)
library(kableExtra)

# data
student_debt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')

# glimpse of the data
student_debt %>%
  head(5) %>%
  knitr::kable(caption = "Top 5 lines of the data") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

Great thing about this data is it's been already cleaned! To know more about the cleaning steps feel free to go to the documentation [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-02-09/readme.md). Here's a short description of the data:

| variable        | class     | description                              |
|-----------------|-----------|------------------------------------------|
| year            | double    | Year of measure                          |
| race            | character | Racial group                             |
| loan\_debt      | double    | Loan debt                                |
| loan\_debt\_pct | double    | Share of families with student loan debt |

: Data Description

## Making an Animated Plot <small>One plot at a time</small>

I will use `gganimate` package to plot the plot. Before using `gganimate` we will first make a plot that we like using `ggplot`, the mighty data viz package from R. In the following code blocks and subsequent paragraphs I will walk you through step by step on how to do that.

### A static line plot

Let's start with a very simple line graph that will show yearly trend of student debt in different races. This plot will work as the skeleton on top of which we'll keep adding new layers and additional bells and whistles to make it nicer and meaningful.

```{r plot1}

student_debt %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race)
  ) +
    geom_line()

```

Some immediate improvement points are: 

* Adding a title and labeling the axis,
* The default values shown in x and y axis are not very meaningful. Breaking down the x and y axis values might make them clearer.


```{r plot2}
student_debt %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race)
  ) +
  geom_line(aes(color = race)) +
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  labs(title="Average Family Student Loan",
         x = NULL, color = NULL, y = "Average Loan Debt")
```

That already looks quite better! 

### A plot with additional insight

We have an unused but very crucial data available to us: `loan_debt_pct`, percent of families with student loan debt. We can add that insight in our plot by using the newly added points!

```{r plot3}

student_debt %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race)
  ) +
  
  geom_line(aes(color = race)) +
  
  geom_point(aes(
    size = loan_debt_pct)
             ) +
  
  
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  labs(title="Average Family Student Loan",
         color = NULL,
         x = NULL, y = "Average Loan Debt",
         subtitle ="Point sizes represent % of families with loans")
```

Now let's make some more improvements:

* The color of the points on different lines represent the corresponding races. Let's make that clear by having them same color as the lines,
* The color more representative to the corresponding races e.g. black for Blacks, white for Whites and brown for Hispanics,
* The legend for races are also not aligned with the sequence of the lines on the graph, let's rearrange the legends too,
* The size legend also needs some reworking. Since it's tough to measure size of circles just from a view, let's take those legends out and rather call out actual percents on the plot.


```{r plot4}
student_debt %>%
  mutate(race = fct_relevel(race, "Black", "White", "Hispanic")) %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race,
        color = race)
  ) +
  
  geom_line(aes()) +
  
  geom_point(aes(
    size = loan_debt_pct),
    show.legend = FALSE) +
  
  geom_text(aes(
    label = paste0(round(loan_debt_pct * 100, 2), "%")
  ), 
  show.legend = FALSE,
  hjust = 1, vjust = 0) + 
  
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#787575", "Hispanic" = "#f5bf42")) + 
                                
  labs(title="Average Family Student Loan",
         color = NULL,
         x = NULL, y = "Average Loan Debt",
         subtitle ="Point sizes represent % of families with loans")
```

### Making it more visually appealing

Our chart is almost ready content wise. But visually it can be much better. The colors are not blending well on the default ggplot background. To make it better I modified `theme_dark()` from ggplot and created a customized these for this plot. Since this tutorial is not about ggplot theme, I'll not discuss about the theme here. That will be part of another tutorial in the future. But in case you are interested to play with ggplot theme here's a good [tutorial](https://rpubs.com/mclaire19/ggplot2-custom-themes) that I found. 

In the iteration of the plot we'll make following changes:

* Change the default theme to dark theme, 
* Adjust line style to match the dark theme,
* To make the percent point texts less cluttered, let's also make theme appear only for the last three years.

```{r theme, include=FALSE}

#Define theme_race() function

theme_race <- function(){ 
    font <- "Georgia"   #assign font family up front
    color <-  "#0a0a0a" #assign black 
    
    theme_minimal() %+replace%    #replace elements we want to change
    
    theme(
      
      #grid elements
      panel.grid.major = element_blank(),    #strip major gridlines
      panel.grid.minor = element_blank(),
      panel.grid.minor.y = element_line(colour = "grey30"),    #strip minor gridlines
      axis.ticks = element_blank(),          #strip axis ticks
      
      #text elements
      text = element_text(colour = "#ffffff",
                          margin = margin(t = 1.5, r = 1, b = 1, l = 1.5, unit = "pt")),
      
      plot.title = element_text(             #title
                   family = font,            #set font family
                   size = 20,                #set font size
                   face = 'bold',            #bold typeface
                   hjust = 0,                #left align
                   vjust = 2,                #raise slightly
                   margin = margin(t = 10, unit = "pt")), 
      
      plot.subtitle = element_text(          #subtitle
                   family = font,            #font family
                   size = 14,                #font size
                   hjust = 0,                #left align
                   vjust = 2,                #raise slightly
                   margin = margin(t = 10, unit = "pt")),              
      
      plot.caption = element_text(           #caption
                   family = font,            #font family
                   size = 9,                 #font size
                   hjust = 1),               #right align
      
      axis.title = element_text(             #axis titles
                   family = font,            #font family
                   size = 10,
                   vjust = 2),               #font size
      
      axis.text = element_text(              #axis text
                   family = font,            #axis famuly
                   size = 9),                #font size
      
      axis.text.x = element_text(            #margin for axis text
                    margin=margin(5, b = 10)),
      
      # adding background color
      panel.background = element_rect(fill = color),
      plot.background = element_rect(fill = color),
      legend.background = element_rect(fill = color),
      legend.key = element_rect(fill = color),
      legend.position = "right",
      legend.direction = "vertical"
    )
}
```

```{r plot5}
student_debt %>%
  mutate(race = fct_relevel(race, "Black", "White", "Hispanic")) %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race,
        color = race)
  ) +
  
  geom_line(aes(),
            size = 1,
            linetype = "dashed") +
  
  geom_point(aes(
    size = loan_debt_pct),
    show.legend = FALSE) +
  
  geom_text(aes(
    label = ifelse(year >= 2010, 
                   paste0(round(loan_debt_pct * 100), "%"),
                   "")
    ), 
  show.legend = FALSE,
  size = 4,
  hjust = 1, vjust = 0) + 
  
  theme_race() + 
  
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#787575", "Hispanic" = "#f5bf42")) + 
                                
  labs(title="Average Family Student Loan",
         color = NULL,
         x = NULL, y = "Average Loan Debt",
         subtitle ="Point sizes represent % of families with loans")
```

### Let's make the plot move!

I think we are at a good enough state to move on to making the plot animated. Animating the plot is pretty easy. We will add ~2 lines of codes to make it.

```{r animation1, cache=TRUE,  message=FALSE, warning=FALSE, fig.height=6, fig.width=10}
student_debt %>%
  mutate(race = fct_relevel(race, "Black", "White", "Hispanic")) %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        group = race,
        color = race)
  ) +
  
  geom_line(aes(),
            size = 1,
            linetype = "dashed") +
  
  geom_point(aes(
    size = loan_debt_pct),
    show.legend = FALSE) +
  
  geom_text(aes(
    label = ifelse(year >= 2010, 
                   paste0(round(loan_debt_pct * 100), "%"),
                   "")
    ), 
  show.legend = FALSE,
  size = 4,
  hjust = 1, vjust = 0) + 
  
  theme_race() + 
  
  transition_reveal(year) +
  
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#787575", "Hispanic" = "#f5bf42")) + 
                                
  labs(title="Average Family Student Loan",
         color = NULL,
         x = NULL, y = "Average Loan Debt",
         subtitle ="Point sizes represent % of families with loans")
```

In our animation there are two areas that need improvement to make the visual more reader friendly:

* The points and texts are disappearing as soon as the new data point is populated. Let's make them stay,
* Let's add a counter on top of the chart that will show the current year as the plot moves,
* And as a final brush up let's also add some more clarity for the labels and add a source credit.

```{r animation2, cache=TRUE,  message=FALSE, warning=FALSE, fig.height=6, fig.width=10}
student_debt %>%
  mutate(race = fct_relevel(race, "Black", "White", "Hispanic")) %>%
  ggplot(
    aes(x = year, 
        y=loan_debt, 
        color = race)
  ) +
  
  geom_line(aes(),
            size = 1,
            linetype = "dashed") +
  
  geom_point(aes(
    size = loan_debt_pct,
    group = seq_along(year)),
    show.legend = FALSE) +
  
  geom_text(aes(
    label = ifelse(year >= 2010, 
                   paste0(round(loan_debt_pct * 100), "%"),
                   ""),
    group = seq_along(year)
    ), 
  show.legend = FALSE,
  size = 4,
  hjust = 1, vjust = 0) + 
  
  theme_race() + 
  
  transition_reveal(as.integer(year)) + # as.integer(year) makes the year showing in subtitle as integer.
  
  scale_x_continuous(breaks = seq(1989, 2016, 3)) + 
    scale_y_continuous(breaks = seq(500, 15000, 1500), 
                       labels = scales::dollar) +
  scale_color_manual(values = c("White" = "#ffffff", "Black" = "#787575", "Hispanic" = "#f5bf42")) + 
                                
  labs(title="Average Student Loan Taken by the US Families",
       x = NULL, color = NULL,
       y = "Average Loan Debt \n($ normalized to 2016 dollars)", 
       caption = "Source: Urban Institute, and the US Census,\n2017 | Arafath Hossain",
       subtitle ="Point sizes represent % of families with loans \nYear: {frame_along}") -> plot
```

## Final Visual

So here's our final visual. 

```{r final_plot, message=FALSE, warning=FALSE}
plot %>% 
  animate(fps = 8, end_pause = 12, height = 6, width = 10, units = "in", res = 150)
```

